Empirical Exploration of the Performance Bottleneck in Graph Neural Network Training

### 2020-08-25 Manuscript/sec2-review-of-graph-neural-network.tex
update

In this section, we formally define the graph neural network (GNN, for short) and briefly survey typical graph neural networks.
We denote a simple graph $\mathcal{G}$ as $\mathcal{G}=(\mathcal{V}, \mathcal{E})$, where $\mathcal{V}$ and $\mathcal{E}$ are the vertex set and the edge set of $\mathcal{G}$, respectively.
Let $n=|\mathcal{V}|$ and $m=|\mathcal{E}|$ as the number of vertices/edges.
We use $v_i$ $(0 \leq i < n)$ to denote a vertex and $e_{i,j}=(v_i, v_j)$ to denote the edge pointing from $v_i$ to $v_j$.
The adjacency set of $v_i$ is $\mathcal{N}(v_i)=\{v|(v_i, v) \in \mathcal{E}\}$.
We denote a \emph{vector} with a bold lower case letter like $\boldsymbol{x}$ and a \emph{matri}x with a bold upper case letter like $\boldsymbol{X}$.

\subsection{General Structure of Graph Neural Networks}

@@

A GNN layer consists of $n$ graph neurons, where $n$ is the number of vertices in $\mathcal{G}$.
Each graph neuron corresponds to a vertex in $\mathcal{G}$.
In the first GNN layer (Layer 0), the graph neuron of the vertex $v_i$ collects input feature vectors of itself and the vertices $\boldsymbol{x}_j$ that are adjacent to $v_i$ in $\mathcal{G}$ (i.e., $v_j \in \mathcal{N}(v_i)$) from the input layer.
After aggregating input feature vectors and applying the non-linear transformation, the graph neuron outputs a hidden feature vector $\boldsymbol{h}^1_i$ for $v_i$.
Take the demo GNN in \figurename~\ref{fig:general_structure_of_gnn}(a) as the example.
Since $\mathcal{N}(v_3) = \{v_1, v_2, v_4, v_5, v_6\}$, the graph neuron of $v_1$ at layer 0 collects the feature vectors \{$\boldsymbol{x}_1$, $\boldsymbol{x}_2$, $\boldsymbol{x}_3$, $\boldsymbol{x}_4$, $\boldsymbol{x}_5$, $\boldsymbol{x}_6$\} from the input layer and outputs $\boldsymbol{h}^1_1$.
Different GNNs mainly differ in the graph neurons that they use.
We elaborate on their details later.

The connection between the input layer and the first GNN layer is determined by the topology of $\mathcal{G}$.
In the traditional neural networks, neurons of neighboring layers are fully connected.
In GNNs, two graph neurons are connected only if their corresponding vertices have an edge between them in $\mathcal{G}$.
Most real-world graphs are very \emph{sparse}, i.e. $|\mathcal{E}| \ll |\mathcal{V}|^2$.

In the next GNN layer (Layer 1), the graph neuron of $v_i$ collects the hidden feature vectors of itself $\boldsymbol{h}^1_i$ and its neighbors ($\boldsymbol{h}^1_j$ with $v_j \in \mathcal{N}(v_i)$) from the \emph{previous} GNN layer.
Based on the collected hidden vectors, the graph neuron in Layer 1 outputs a new hidden feature vector $\boldsymbol{h}^2_i$ for $v_i$.
Though there are only two GNN layers in \figurename~\ref{fig:general_structure_of_gnn}, a GNN allows us to stack more GNN layers to support deeper graph analysis.
%In \figurename~\ref{fig:general_structure_of_gnn}, we shows a GNN with two layers.

Assume there are $L$ GNN layers.
The last GNN layer (Layer $L-1$) outputs a hidden feature vector $\boldsymbol{h}^{L}_i$ for every vertex $v_i$.
As an embedding vector, $\boldsymbol{h}^L_i$ encodes the knowledge learned from the input layer and all the previous GNN layers.
Since $\boldsymbol{h}^L_i$ is affected by $v_i$ and the vertices in the $L$-hop neighborhood of $v_i$, analyzing a graph with a \emph{deeper} GNN means analyzing each vertex with a \emph{wider} scope.

The hidden feature vectors $\boldsymbol{h}^L_i$ of the last GNN layer are fed to the prediction layer to generate the output of the whole GNN.
The prediction layer is a standard neural network.
The structure of the prediction layer depends on the prediction task of the GNN.
Take the node classification task as the example, as shown in \figurename~\ref{fig:general_structure_of_gnn}.
The node classification predicts a label for every vertex in $\mathcal{G}$.
In this case, the prediction layer can be a simple softmax layer with $\boldsymbol{h}^L_i$ as the input and a vector of probabilities as the output.
If the prediction task is edge prediction, the hidden feature vectors of two vertices are concatenated and fed into a softmax layer.
If we need to predict a label for the whole graph, a pooling (max/mean/...) layer is added to generate an embedding vector for the whole graph and the embedding vector is used to produce the final prediction.

Supporting end-to-end training is a prominent advantage of GNN, compared with other graph-based machine learning methods.
We can calculate the gradients of the loss function on the model parameters from the prediction layer directly.
With the help of the backpropagation technique, the gradient is propagated from the prediction layer back to the previous GNN layers layer by layer.
The model parameters are updated with a gradient descent optimizer like Adam.
Except for the input feature vector, there is no need to conduct handworked feature extraction.
In a fully parameterized way, the GNN automatically extracts an embedding vector for each vertex from its $L$-hop neighborhood.
The parameters are tuned according to the specific prediction task, leading to high prediction accuracy.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.5\columnwidth]{figs/illustration/GNN_Unit.pdf}
    \caption{Graph neuron of $v_3$ at the GNN layer $l$ with the graph $\mathcal{G}$ in \figurename~\ref{fig:general_structure_of_gnn}(b). $\phi$/$\Sigma$/$\gamma$ are the message/aggregation/vertex update functions in the message-passing model, respecitvely.}
    \label{fig:graph_neuron_structure}
\end{figure}

We follow the message-passing model \cite{gilmer_messgae_passing} to formally define a graph neuron.
The message-passing model is widely used in the cutting-edge GNN training systems like PyTorch Geometric (PyG) \cite{PyG} and Deep Graph Library (DGL) \cite{DGL}.
\figurename~\ref{fig:graph_neuron_structure} shows the structure of a graph neuron in the message-passing model.
A graph neuron collects messages from its neighbor graph neurons and outputs a hidden feature vector according to the aggregated messages.
Graph neurons at layer $l$ are made of three \emph{differentiable} functions: $\phi^l$, $\Sigma^l$ and $\gamma^l$.
The graph neuron calculates the output hidden vector $\boldsymbol{h}^{l+1}_i$ by
$$
    \boldsymbol{h}^{l+1}_i = \gamma^l(\boldsymbol{h}^l_i, \mathlarger{\Sigma}^l_{v_j \in \mathcal{N}(v_i)}{\phi^l(\boldsymbol{h}^l_i, \boldsymbol{h}^l_j,   \boldsymbol{e}_{j,i})}).
$$

$\phi^l$ is the \emph{message} function.
For every incident edge $(v_j, v_i)$ of $v_i$, $\phi$ receives the output hidden feature vectors $\boldsymbol{h}^l_i$ and $\boldsymbol{h}^l_j$ of the previous GNN layer and the edge feature vector $\boldsymbol{e}_{j,i}$ as the input.
$\phi^l$ emits a message vector $\boldsymbol{m}^l_{j,i}$ for every edge $(v_j, v_i)$ at layer $l$, i.e., $\boldsymbol{m}^l_{j,i}=\phi^l(\boldsymbol{h}^l_i, \boldsymbol{h}^l_j, \boldsymbol{e}_{j,i})$.
For $v_i$, the message vectors $\boldsymbol{m}^l_{x,j}$ of its adjacent edges ($v_x \in \mathcal{N}(v_i)$) are aggregated by the \emph{aggregation} function $\Sigma^l$ to produce an aggregated vector $\boldsymbol{s}^l_i$, i.e., $\boldsymbol{s}^l_i=\mathlarger{\Sigma}^l_{v_j \in \mathcal{N}(v_i)}\boldsymbol{m}^l_{j,i}$.
$v_i$'s aggregated vector $\boldsymbol{s}^l_i$ and its hidden vector $\boldsymbol{h}^l_i$ from the previous GNN layer are fed into the \emph{vertex update} function $\gamma^l$ to calculate the output hidden vector $\boldsymbol{h}^{l+1}_i$ of the current layer $l$, i.e., $\boldsymbol{h}^{l+1}_i = \gamma^l(\boldsymbol{h}^l_i, \boldsymbol{s}^l_i)$
The end-to-end training requires $\phi^l$ and $\gamma^l$ (like multi-layer perceptrons and GRU) and $\Sigma_l$ (like mean, sum, element-wise min/max) are \emph{differentiable} to make the whole GNN differentiable.

Different GNNs adopt different kinds of graph neurons and have different definitions of the three functions.
$\phi$ and $\Sigma$ are the \emph{edge calculation} functions.
They are conducted over every edge in $\mathcal{G}$.
$\gamma$ is the \emph{vertex calculation} function.
It is conducted over every vertex in $\mathcal{G}$.
\tablename~\ref{tab:gnn_overview_edge} and \tablename~\ref{tab:gnn_overview_vertex} list the edge functions and the vertex functions of typical GNNs, respectively.
For ChebNet, we report its GNN sub-layer in the tables.
A ChebNet layer consists of $K$ GNN sub-layers and a summation layer:
$\boldsymbol{H}^{l+1} = \sum_{k=1}^K{\boldsymbol{Z}^{(k)} \textcolor{blue}{\boldsymbol{W}^{(k)}}}$ with the GNN sub-layers $\boldsymbol{Z}^{(1)}=\boldsymbol{H}^l$, $\boldsymbol{Z}^{(2)}=\hat{\boldsymbol{L}}\boldsymbol{H}^l$, and $\boldsymbol{Z}^{(k)}=2\hat{\boldsymbol{L}}\boldsymbol{Z}^{(k-1)} - \boldsymbol{Z}^{(k-2)}$, where $\boldsymbol{H}^l$ is the matrix of output hidden feature vectors of the layer $l-1$.
For GAT, a GAT layer consists of two sub-layers and it conducts part of the vertex calculation before the two sub-layers.
For GaAN, a GaAN layer consists of four sub-layers: the first sub-layer calculates the summation $\boldsymbol{m}_{j, i, (k)}^{l,0}  = \exp((\textcolor{blue}{W^{l, k}_{xa}} \boldsymbol{h}_j^l + \textcolor{blue}{b^{l, k}_{xa}})^T (\textcolor{blue}{\boldsymbol{W}^{l, k}_{ya}} \boldsymbol{h}_i^l + \textcolor{blue}{\boldsymbol{b}^{l,k}_{ya}}))$, and the other three layers calculate $\boldsymbol{m}^l_{j,i,1}$/$\boldsymbol{m}^l_{j,i,2}$/$\boldsymbol{m}^l_{j,i,3}$.

@@

            $\boldsymbol{m}_{j, i}^l =  \delta(\textcolor{blue}{\boldsymbol{W}^l_{pool}} \boldsymbol{h}_j^l + \textcolor{blue}{b}^l)$ &

@@

                         & {\boldsymbol{z}}_i^l = \delta ( \textcolor{blue}{\boldsymbol{W}}^z \boldsymbol{s}_i^l + \textcolor{blue}{\boldsymbol{b}}^{sz} + \textcolor{blue}{\boldsymbol{U}}^z \boldsymbol{h}_i^{l} + \textcolor{blue}{\boldsymbol{b}}^{hz})                    \\
                         & \boldsymbol{r}_i^l = \delta ( \textcolor{blue}{\boldsymbol{W}}^r \boldsymbol{s}_i^l+ \textcolor{blue}{\boldsymbol{b}}^{sr} +\textcolor{blue}{\boldsymbol{U}}^r \boldsymbol{h}_i^{l} + \textcolor{blue}{\boldsymbol{b}}^{hr})                        \\
                         & \boldsymbol{h}_i^{l+1} = tanh ( \textcolor{blue}{\boldsymbol{W}} \boldsymbol{s}_i^l + \textcolor{blue}{\boldsymbol{b}}^s + \textcolor{blue}{\boldsymbol{U}} ( \boldsymbol{r}_i^l \odot \boldsymbol{h}_i^{l}) + \textcolor{blue}{\boldsymbol{b}}^h)) \\

@@

                         & \text{Sub-layer 1, 2, 3:}                                                                                                                                                                      \\

@@

In \tablename~\ref{tab:gnn_overview_edge} and \tablename~\ref{tab:gnn_overview_vertex}, we list the edge and vertex calculation complexity, respectively.
The time complexity of a graph neuron is affected by the dimensions of the input/output hidden vectors $d_{in}$ and $d_{out}$ and the dimensions of the model parameters (like the number of heads $K$ in GAT and the dimensions of the view vectors $d_a$/$d_v$ in GaAN).

@@

\textbf{GCN} \cite{kipf2017_gcn} (low vertex \& edge calculation complexity): Graph convolution network (GCN) is the first-order approximation of the spectral-based graph convolutions.
%By combining the advantages of both the spectral-based and the spatial-based convolution, it can learn the representation of the local topological structure and vertex features efficiently.
It has only one parameter to learn at each layer, i.e. the weight matrix $\boldsymbol{W}^l$ in $\gamma$.
A GCN graph neuron can be expressed as $\boldsymbol{h}^{l+1}_i = \boldsymbol{W}^l\sum_{v_j \in \mathcal{N}(v_i)}{e_{j,i}\boldsymbol{h}^l_j}$, where $e_{j,i}$ is the normalized weight of the edge $(v_j, v_i)$.
%In the matrix view, $\boldsymbol{H}^{l+1} = (\boldsymbol{A}\boldsymbol{H}^l)\boldsymbol{W}^l$.
According to the associative law of the matrix multiplication, $\boldsymbol{h}^{l+1}_i = \sum_{v_j \in \mathcal{N}(v_i)}{e_{j,i}\boldsymbol{W}^l\boldsymbol{h}^l_j}$.
Since the dimension of $\boldsymbol{h}^{l+1}_i$ is usually smaller than $\boldsymbol{h}^l_i$ in practical GCNs, the implementation of GCN in PyTorch Geometric chooses to first conduct the vertex calculation $\hat{\boldsymbol{h}}^l_j = \boldsymbol{W}^l\boldsymbol{h}^l_j$ for each vertex $v_j$ and then conduct the edge calculation $\boldsymbol{h}^{l+1}_i=\sum_{v_j\in\mathcal{N}(v_i)}{e_{j,i}\hat{\boldsymbol{h}}^l_j}$.
As $\hat{\boldsymbol{h}}^l_j$ has the same dimension as $\boldsymbol{h}^{l+1}_i$, the implementation significantly reduces the  computation cost of the edge calculation.

\textbf{GGNN} (high vertex \& low edge calculation complexity):
GGNN introduces the gated recurrent unit (GRU) into the graph neural networks.
The vertex update function $\phi$ of GGNN is a modified GRU unit that has 12 model parameters to learn, having high computational complexity.
To lower the training cost, all GNN layers share the same group of parameters in GGNN.
GGNN further requires the dimension of $\boldsymbol{h}^{l+1}$ is equal to the dimension of $\boldsymbol{h}^l$.
Since the message function $\phi$ only uses the hidden feature vector $\boldsymbol{h}^l_j$ of the source vertex $v_j$ of an edge $(v_j, v_i)$, in the implementation, GGNN conducts the pre-processing vertex calculation $\hat{\boldsymbol{h}}^l_i=\boldsymbol{W}\boldsymbol{h}^l_i$ for every vertex $v_i$ before the message-passing.
The message function $\phi$ directly uses $\hat{\boldsymbol{h}}^l_j$ as the message vector for the edge $(v_j, v_i)$.
In this way, GGNN further reduces the time complexity of the edge calculation to $O(1)$ without increasing the time complexity of the vertex calculation.

\textbf{GAT} (low vertex \& high edge calculation complexity):
GAT introduces the attention and multi-head mechanism into the graph neural networks.
The $K$ heads generate $K$ independent views for an edge, where $K$ is a hyper-parameter.
The views of $K$ heads can be merged by concatenating or by averaging.
For concatenating, the dimension of the hidden feature vector of each head $d_{head}$ is $d_{out}/K$.
For averaging, $d_{head}$ is $d_{out}$, multiplying the complexity by $K$.
Each GAT layer consists of a preprocessing step and two sub-layers.
In the preprocessing step, GAT calculates the attention vectors $\hat{\boldsymbol{h}}^{l}_{i,(k)}$ for every vertex $v_i$ in every head $k$.
The first sub-layer uses the attention vectors to calculate the attention weights of every edge in every head $\boldsymbol{m}^{l,0}_{j,i,(k)}$.
After the message-passing, the first-layer gets the summation of attention weights of every vertex in every head $\hat{\boldsymbol{a}}^{l,0}_{i,(k)}$.
The second sub-layer gets the normalized attention weights $\alpha_{j, i, (k)}$ for every edge in every head and aggregates the hidden feature vectors with the normalized weights in the message-passing.
GAT uses the aggregated vector $\boldsymbol{s}^{l,1}_i$ of the second sub-layer as the output hidden vector directly.

>>>>

In this section, we formally define the graph neural networks and briefly survey typical graph neural networks.
We denote a simple graph $\mathcal{G}$ as $\mathcal{G}=(\mathcal{V}, \mathcal{E})$, where $\mathcal{V}$ and $\mathcal{E}$ are the vertex set and the edge set of $\mathcal{G}$, respectively.
Let $n=|\mathcal{V}|$ and $m=|\mathcal{E}|$ as the number of vertices/edges.
We use $v_i$ $(0 \leq i < n)$ to denote a vertex and $e_{i,j}=(v_i, v_j)$ to denote the edge pointing from $v_i$ to $v_j$.
The adjacency set of $v_i$ is $\mathcal{N}(v_i)=\{v|(v_i, v) \in \mathcal{E}\}$.
We denote a \emph{vector} with a bold lower case letter like $\boldsymbol{x}$ and a \emph{matri}x with a bold upper case letter like $\boldsymbol{X}$.

\subsection{Structure of Graph Neural Networks}

@@

A GNN layer consists of $n$ graph neurons, where $n$ is the number of vertices in $\mathcal{G}$.
Each graph neuron corresponds to a vertex in $\mathcal{G}$.
In the first GNN layer (Layer 0), the graph neuron of the vertex $v_i$ collects input feature vectors of itself and the vertices $\boldsymbol{x}_j$ that are adjacent to $v_i$ in $\mathcal{G}$ (i.e., $v_j \in \mathcal{N}(v_i)$) from the input layer.
The graph neuron aggregates input feature vectors, apply non-linear transformation, and outputs a hidden feature vector $\boldsymbol{h}^1_i$ for $v_i$.
Take the demo GNN in \figurename~\ref{fig:general_structure_of_gnn}(a) as the example.
Since $\mathcal{N}(v_3) = \{v_1, v_2, v_4, v_5, v_6\}$, the graph neuron of $v_3$ at the GNN layer 0 collects the feature vectors \{$\boldsymbol{x}_1$, $\boldsymbol{x}_2$, $\boldsymbol{x}_3$, $\boldsymbol{x}_4$, $\boldsymbol{x}_5$, $\boldsymbol{x}_6$\} from the input layer and outputs $\boldsymbol{h}^1_3$.
Different GNNs mainly differ in the graph neurons that they use.
We elaborate on their details later.

Different from the traditional neural networks that layers are fully connected, the layers are sparsely connected in GNNs. 
The connection between the input layer and the first GNN layer is determined by the topology of $\mathcal{G}$.
Two graph neurons are connected only if their corresponding vertices have an edge between them in $\mathcal{G}$.
Most real-world graphs are very \emph{sparse}, i.e. $|\mathcal{E}| \ll |\mathcal{V}|^2$.

In the next GNN layer (GNN Layer 1), the graph neuron of $v_i$ collects the hidden feature vectors of itself $\boldsymbol{h}^1_i$ and its neighbors ($\boldsymbol{h}^1_j$ with $v_j \in \mathcal{N}(v_i)$) from the \emph{previous} GNN layer.
Based on the collected hidden vectors, the graph neuron in Layer 1 outputs a new hidden feature vector $\boldsymbol{h}^2_i$ for $v_i$.
A GNN allows stacking more GNN layers to support deeper graph analysis.
Assume there are $L$ GNN layers in traditional.
The last GNN layer (GNN Layer $L-1$) outputs a hidden feature vector $\boldsymbol{h}^{L}_i$ for every vertex $v_i$.
As an embedding vector, $\boldsymbol{h}^L_i$ encodes the knowledge learned from the input layer and all the previous GNN layers.
Since $\boldsymbol{h}^L_i$ is affected by $v_i$ and the vertices in the $L$-hop neighborhood of $v_i$, analyzing a graph with a \emph{deeper} GNN means analyzing each vertex with a \emph{wider} scope.

The hidden feature vectors $\boldsymbol{h}^L_i$ of the last GNN layer are fed to the prediction layer to generate the output of the whole GNN.
The prediction layer is a standard neural network.
The structure of the prediction layer depends on the prediction task of the GNN.
Take the node classification task in \figurename~\ref{fig:general_structure_of_gnn} as the example.
The node classification predicts a label for every vertex in $\mathcal{G}$.
In this case, the prediction layer can be a simple softmax layer with $\boldsymbol{h}^L_i$ as the input and a vector of probabilities as the output.
If the prediction task is edge prediction, the hidden feature vectors of two vertices are concatenated and fed into a softmax layer.
If we need to predict a label for the whole graph, a pooling (max/mean/...) layer is added to generate an embedding vector for the whole graph and the embedding vector is used to produce the final prediction.

Supporting end-to-end training is a prominent advantage of GNN, compared with traditional graph-based machine learning methods.
The traditional methods need to construct feature vectors manually or from the embedding methods like DeepWalk \cite{bryan2014_deepwalk} and node2vec \cite{aditya2016_node2vec}.
The feature vector generation is independent from the downstream prediction task. 
Thus, the vectors have to be adjusted for every prediction task individually.
In GNNs, the gradients are propagated from the prediction layer back to the previous GNN layers layer by layer. 
The model parameters in the GNN layers are updated based on the feedbacks from the downstream prediction task. 
In a fully parameterized way, the GNN automatically extracts an embedding vector for each vertex from its $L$-hop neighborhood, tuned according to the specific prediction task.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.5\columnwidth]{figs/illustration/GNN_Unit.pdf}
    \caption{Graph neuron of $v_3$ at the GNN layer $l$ with the graph $\mathcal{G}$ in \figurename~\ref{fig:general_structure_of_gnn}(b). $\phi$/$\Sigma$/$\gamma$ are the messaging/aggregation/vertex updating functions in the message-passing model, respecitvely.}
    \label{fig:graph_neuron_structure}
\end{figure}

We follow the message-passing model \cite{gilmer_messgae_passing} to formally define a graph neuron.
The message-passing model is widely used in the cutting-edge GNN libraries like PyTorch Geometric (PyG) \cite{PyG} and Deep Graph Library (DGL) \cite{DGL}.
\figurename~\ref{fig:graph_neuron_structure} shows the structure of a graph neuron in the message-passing model.
Graph neurons at layer $l$ are made of three \emph{differentiable} functions: $\phi^l$, $\Sigma^l$ and $\gamma^l$.
The graph neuron calculates the output hidden vector $\boldsymbol{h}^{l+1}_i$ by
$$
    \boldsymbol{h}^{l+1}_i = \gamma^l(\boldsymbol{h}^l_i, \mathlarger{\Sigma}^l_{v_j \in \mathcal{N}(v_i)}{\phi^l(\boldsymbol{h}^l_i, \boldsymbol{h}^l_j,   \boldsymbol{e}_{j,i})}).
$$

$\phi^l$ is the \emph{messaging} function.
For every incident edge $(v_j, v_i)$ of $v_i$, $\phi$ receives the output hidden feature vectors $\boldsymbol{h}^l_i$ and $\boldsymbol{h}^l_j$ from the previous GNN layer and the edge feature vector $\boldsymbol{e}_{j,i}$ as the input.
$\phi^l$ emits a message vector $\boldsymbol{m}^l_{j,i}$ for every edge $(v_j, v_i)$ at layer $l$, i.e., $\boldsymbol{m}^l_{j,i}=\phi^l(\boldsymbol{h}^l_i, \boldsymbol{h}^l_j, \boldsymbol{e}_{j,i})$.
For every $v_i$, the message vectors $\boldsymbol{m}^l_{j,i}$ of its adjacent edges ($v_j \in \mathcal{N}(v_i)$) are aggregated by the \emph{aggregation} function $\Sigma^l$ to produce an aggregated vector $\boldsymbol{s}^l_i$, i.e., $\boldsymbol{s}^l_i=\mathlarger{\Sigma}^l_{v_j \in \mathcal{N}(v_i)}\boldsymbol{m}^l_{j,i}$.
$v_i$'s aggregated vector $\boldsymbol{s}^l_i$ and its hidden vector $\boldsymbol{h}^l_i$ from the previous GNN layer are fed into the \emph{vertex updating} function $\gamma^l$ to calculate the output hidden vector $\boldsymbol{h}^{l+1}_i$ of the current layer $l$, i.e., $\boldsymbol{h}^{l+1}_i = \gamma^l(\boldsymbol{h}^l_i, \boldsymbol{s}^l_i)$
The end-to-end training requires $\phi^l$ and $\gamma^l$ (like multi-layer perceptrons and GRU) and $\Sigma_l$ (like mean, sum, element-wise min/max) are \emph{differentiable} to make the whole GNN differentiable.

Different GNNs have different definitions of the three functions.
We regard $\phi$ and $\Sigma$ as the \emph{edge calculation} functions, since they are conducted over every edge in $\mathcal{G}$.
We regard $\gamma$ as the \emph{vertex calculation} function, as it is conducted over every vertex in $\mathcal{G}$.
\tablename~\ref{tab:gnn_overview_edge} and \tablename~\ref{tab:gnn_overview_vertex} list the edge functions and the vertex functions of typical GNNs, respectively.
For ChebNet, we report its GNN sub-layer in the tables.
A ChebNet layer consists of $K$ GNN sub-layers and a summation layer:
$\boldsymbol{H}^{l+1} = \sum_{k=1}^K{\boldsymbol{Z}^{(k)} \textcolor{blue}{\boldsymbol{W}^{(k)}}}$ with the GNN sub-layers $\boldsymbol{Z}^{(1)}=\boldsymbol{H}^l$, $\boldsymbol{Z}^{(2)}=\hat{\boldsymbol{L}}\boldsymbol{H}^l$, and $\boldsymbol{Z}^{(k)}=2\hat{\boldsymbol{L}}\boldsymbol{Z}^{(k-1)} - \boldsymbol{Z}^{(k-2)}$, where $\boldsymbol{H}^l$ is the matrix of output hidden feature vectors of the layer $l-1$.
For GAT, a GAT layer consists of two sub-layers and it conducts part of the vertex calculation before the two sub-layers.
For GaAN, a GaAN layer consists of four sub-layers: the first sub-layer calculates the summation $\boldsymbol{m}_{j, i, (k)}^{l,0}  = \exp((\textcolor{blue}{W^{l, k}_{xa}} \boldsymbol{h}_j^l + \textcolor{blue}{b^{l, k}_{xa}})^T (\textcolor{blue}{\boldsymbol{W}^{l, k}_{ya}} \boldsymbol{h}_i^l + \textcolor{blue}{\boldsymbol{b}^{l,k}_{ya}}))$, and the other three sub-layers calculate $\boldsymbol{m}^l_{j,i,1}$/$\boldsymbol{m}^l_{j,i,2}$/$\boldsymbol{m}^l_{j,i,3}$.

@@

            $\boldsymbol{m}_{j, i}^l =  \delta(\textcolor{blue}{\boldsymbol{W}^l_{pool}} \boldsymbol{h}_j^l + \textcolor{blue}{b^l})$ &

@@

                         & {\boldsymbol{z}}_i^l = \delta ( \textcolor{blue}{\boldsymbol{W}^z} \boldsymbol{s}_i^l + \textcolor{blue}{\boldsymbol{b}^{sz}} + \textcolor{blue}{\boldsymbol{U}^z} \boldsymbol{h}_i^{l} + \textcolor{blue}{\boldsymbol{b}^{hz}})                    \\
                         & \boldsymbol{r}_i^l = \delta ( \textcolor{blue}{\boldsymbol{W}^r} \boldsymbol{s}_i^l+ \textcolor{blue}{\boldsymbol{b}^{sr}} +\textcolor{blue}{\boldsymbol{U}^r} \boldsymbol{h}_i^{l} + \textcolor{blue}{\boldsymbol{b}^{hr}})                        \\
                         & \boldsymbol{h}_i^{l+1} = tanh ( \textcolor{blue}{\boldsymbol{W}} \boldsymbol{s}_i^l + \textcolor{blue}{\boldsymbol{b}^s} + \textcolor{blue}{\boldsymbol{U}} ( \boldsymbol{r}_i^l \odot \boldsymbol{h}_i^{l}) + \textcolor{blue}{\boldsymbol{b}^h})) \\

@@

                         & \text{After sub-layer 0,1,2,3:}                                                                                                                                                                      \\

@@

In \tablename~\ref{tab:gnn_overview_edge} and \tablename~\ref{tab:gnn_overview_vertex}, we list the edge and vertex calculation complexity, respectively.
The time complexity of a graph neuron is affected by the dimensions of the input/output hidden vectors $d_{in}$ and $d_{out}$ and the dimensions of the model parameters (like the number of heads $K$ in GAT and the dimensions of the view vectors $d_a$/$d_v$/$d_m$ in GaAN).

@@

\textbf{GCN} \cite{kipf2017_gcn} (low vertex \& edge calculation complexity): Graph convolution network (GCN \cite{kipf2017_gcn}) introduces the first-order approximation of the spectral-based graph convolutions.
%By combining the advantages of both the spectral-based and the spatial-based convolution, it can learn the representation of the local topological structure and vertex features efficiently.
It has only one parameter to learn at each layer, i.e. the weight matrix $\boldsymbol{W}^l$ in $\gamma$.
A GCN graph neuron can be expressed as $\boldsymbol{h}^{l+1}_i = \boldsymbol{W}^l\sum_{v_j \in \mathcal{N}(v_i)}{e_{j,i}\boldsymbol{h}^l_j}$, where $e_{j,i}$ is the normalized weight of the edge $(v_j, v_i)$.
%In the matrix view, $\boldsymbol{H}^{l+1} = (\boldsymbol{A}\boldsymbol{H}^l)\boldsymbol{W}^l$.
According to the associative law of the matrix multiplication, $\boldsymbol{h}^{l+1}_i = \sum_{v_j \in \mathcal{N}(v_i)}{e_{j,i}\boldsymbol{W}^l\boldsymbol{h}^l_j}$.
Since the dimension of $\boldsymbol{h}^{l+1}_i$ is usually smaller than $\boldsymbol{h}^l_i$ in practical GCNs, the implementation of GCN in PyTorch Geometric chooses to first conduct the vertex calculation $\hat{\boldsymbol{h}}^l_j = \boldsymbol{W}^l\boldsymbol{h}^l_j$ for each vertex $v_j$ and then conduct the edge calculation $\boldsymbol{h}^{l+1}_i=\sum_{v_j\in\mathcal{N}(v_i)}{e_{j,i}\hat{\boldsymbol{h}}^l_j}$.
As $\hat{\boldsymbol{h}}^l_j$ has the same dimension as $\boldsymbol{h}^{l+1}_i$, the implementation significantly reduces the  computation cost of the edge calculation.

\textbf{GGNN} (high vertex \& low edge calculation complexity):
GGNN \cite{li2015_ggnn} introduces the gated recurrent unit (GRU) into the graph neural networks.
The vertex updating function $\phi$ of GGNN is a modified GRU unit that has 12 model parameters to learn, having high computational complexity.
To lower the training cost, all GNN layers share the same group of parameters in GGNN.
GGNN further requires the dimension of $\boldsymbol{h}^{l+1}$ is equal to the dimension of $\boldsymbol{h}^l$.
Since the messaging function $\phi$ only uses the hidden feature vector $\boldsymbol{h}^l_j$ of the source vertex $v_j$ of an edge $(v_j, v_i)$. In the implementation of PyG, GGNN conducts the pre-processing vertex calculation $\hat{\boldsymbol{h}}^l_i=\boldsymbol{W}\boldsymbol{h}^l_i$ for every vertex $v_i$ before the message-passing.
The messaging function $\phi$ directly uses $\hat{\boldsymbol{h}}^l_j$ as the message vector for the edge $(v_j, v_i)$.
In this way, GGNN further reduces the time complexity of the edge calculation to $O(1)$ without increasing the time complexity of the vertex calculation.

\textbf{GAT} (low vertex \& high edge calculation complexity):
GAT \cite{huang2018_gat} introduces the attention and multi-head mechanism into the graph neural networks.
The $K$ heads generate $K$ independent views for an edge, where $K$ is a hyper-parameter.
The views of $K$ heads can be merged by concatenating or by averaging.
For concatenating, the dimension of the hidden feature vector of each head $d_{head}$ is $d_{out}/K$.
For averaging, $d_{head}$ is $d_{out}$, multiplying the complexity by $K$.
Each GAT layer consists of a preprocessing step and two sub-layers.
In the preprocessing step, GAT calculates the attention vectors $\hat{\boldsymbol{h}}^{l}_{i,(k)}$ for every vertex $v_i$ in every head $k$.
The first sub-layer uses the attention vectors to calculate the attention weights of every edge in every head $\boldsymbol{m}^{l,0}_{j,i,(k)}$.
After aggregation, the first sub-layer gets the summation of attention weights of every vertex in every head $\hat{\boldsymbol{a}}^{l,0}_{i,(k)}$.
The second sub-layer calculates the normalized attention weights $\alpha_{j, i, (k)}$ for every edge in every head and aggregates the hidden feature vectors with the normalized weights in the message-passing.
GAT uses the aggregated vector $\boldsymbol{s}^{l,1}_i$ of the second sub-layer as the output hidden vector directly.

\textbf{GaAN} (high vertex \& edge calculation complexity):
Based on the multi-head mechanism, GaAN \cite{zhang2018_gaan} introduces a convolutional subnetwork to control the weight of each head.
Each GaAN layer consists of four sub-layers. The first two sub-layers are similar to GAT, and the last two sub-layers build the convolutional subnetwork.
The first sub-layer aims to get the summation of attention weights of every vertex in each head $\boldsymbol{\hat{a}}^{l,0}_{j, i, (k)}$.
The second sub-layer aggregates the feature vectors with the normalized attention weights $\boldsymbol{\hat{a}}^{l,0}_{i,(k)}$.
The third and fourth sub-layers aggregate feature vectors with element-wise max and element-wise mean operators separately.
After the four sub-layers, the vertex updating funciton uses the aggregated vectors of last three sub-layers $\boldsymbol{s}^{l, 1}_i, \boldsymbol{s}^{l, 2}_i, \boldsymbol{s}^{l, 3}_i$ to generate the output hidden vector $\boldsymbol{h}^{l+1}_i$
